#### Numerical Summaries  

To create numerical summaries in R, you first need to determine the type of data you have.  Remember there are two major categories:

+ Categorical (Qualitative) variable - entries are a label or attribute   
+ Numerical (Quantitative) variable - entries are a numerical value where math can be performed

The main goal is to describe the distribution or pattern and frequency with which you observe the variable.  For categorical variables, we used contingency tables to describe the count in each category.  In R, this is done using the `table` function.  

We'll recreate the summaries of the titanic dataset (available in the file `titanic.csv` in the datasets folder).  

```{r,echo=TRUE}
titanicData <- read_csv("../../datasets/titanic.csv")
titanicData
```

Recall the three variables we created tables for were:

+ embarked (where journey started)  
+ survived (survive or not)    
+ sex (Male or Female)  

To create the **one-way contingency tables**, we simply pass the column of interest to the `table` function.  

```{r}
table(titanicData$embarked)
table(titanicData$survived)
table(titanicData$sex)
```

To create a **two-way contingency table**, we can pass the columns of interest from the dataset.  

```{r}
table(titanicData$survived, titanicData$sex)
table(titanicData$survived, titanicData$embarked)
table(titanicData$sex, titanicData$embarked)
```

This same process can continue to create an **n-way contingency table**.

```{r}
table(titanicData$sex, titanicData$embarked, titanicData$survived)
```

These table objects in R are actually arrays - the extension of a matrix to more than two dimensions. 

```{r}
tab <- table(titanicData$sex, titanicData$embarked, titanicData$survived)
str(tab)
```

You can obtain **conditional** bivariate info by supplying an index.

```{r}
#returns embarked vs survived table for females
tab[1, , ]
#returns embarked vs survived table for males
tab[2, , ]
#returns survived vs sex table for embarked "C"
tab[, 1, ]
```

We can also obtain **conditional** univariate info too!

```{r}
#Survived status for males that embarked at "Q"
tab[2, 2, ]
```

For numerical variables, recall that we want to describe the shape, center, and spread of the variable's distribution.  

Recall the carbon dioxide (CO2) uptake data set  

+ Response recorded: `uptake` CO2 uptake rates in grass plants  
+ Environment manipulated: `Treatment` - chilled/nonchilled  
+ Ambient CO2 specified and measured: `conc`  

```{r}
CO2 <- tbl_df(CO2)
CO2
```

To investigate the center we looked at the mean, trimmed mean, and the median.  The `mean` and `median` functions handle these.  You simply pass the column of interest.  To obtain a trimmed mean, you can use the `trim` argument.  

```{r}
mean(CO2$uptake)
#5% trimmed mean
mean(CO2$uptake, trim = 0.05) 
median(CO2$uptake)
```

To obtain measures of spread we can use `var` and `sd`.  Similarly, we pass the column of interest to these functions.  

```{r}
var(CO2$uptake)
sd(CO2$uptake)
```

Quantiles can be found using the `quantile` function.  You pass the column of interest and a vector of probabilities corresponding to the quantiles of interest.  

```{r}
quantile(CO2$uptake, probs = c(0.1, 0.2))
```

The IQR can be found by passing the `IQR` function the column of interest.

```{r}
IQR(CO2$uptake)
```

R has some nice functions that return multiple summaries.  For instance, the `summary` function!

```{r}
#quartiles and mean
summary(CO2$uptake)
```

The `psych` package has a very nice function called `describe` that provides many summaries about your numerical variables in the dataset as well.  Remember you need to install a package if you've never used it and then read it in to your current R session using the `library` or `require` function.  

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(psych)
```

```{r}
describe(titanicData)
describe(CO2)
```

Recall, if you had multiple numerical variables we often want to measure the linear relationship.  This is done via covariance and correlation.  If you pass the `var` or `cov` function a data frame with multiple numeric columns, you'll obtain the variance-covariance matrix.  The diagonals represent the variance and the off-diagonals the covariances.  The `cor` function similarly gives the correlation matrix.

```{r}
var(select(CO2, conc, uptake))
cov(select(CO2, conc, uptake))
```

Of course, often summaries by a subgroup of the data is of interest.  We looked at finding summaries of `uptake` for each `Treatment` level.  `dplyr` is easy to use for this purpose but can only use functions that return a single value.  The idea is:

- Use `group_by` to create subgroups associated with the data frame  
- Use `summarize` to create basic summaries for each subgroup  

For example: 

```{r}
CO2 %>% group_by(Treatment) %>% 
	summarise(avg = mean(uptake), med = median(uptake), var = var(uptake))
```

To find the summary for different settings of two variables, say **Treatment** and **Concentration**, simply pass more than one categorical variable to `group_by`.  

```{r}
CO2 %>% group_by(Treatment, conc) %>% 
		summarise(avg = mean(uptake), med = median(uptake), var = var(uptake))
```

An alternative to using `dplyr` is to use the built-in `aggregate` function.  This function is more flexible and can be used with functions that return more than one value like `summary`.  Basic use gives a response (`x`) and a `list` of variables to group `by`.

```{r}
aggregate(x = CO2$uptake, by = list(CO2$Treatment), FUN = summary)
```

Alternatively, `formula` notation can be used.  We can supply `uptake ~ Treatment` and the data object.  The idea of formula notation is essentially a shorthand for modeling the left-hand side (uptake here) by the right-hand side (Treatment here).  In this case, this implies finding `uptake` values returned by `FUN` for the different levels of `Treatment`. 

```{r}
aggregate(uptake ~ Treatment, data = CO2, FUN = summary)
```

A second variable can be given on the right hand side.  The summaries will then be found for every combination of the values of the variables on the right hand side.  Since `conc` is only observed at a few levels, this is reasonable here.  

```{r}
aggregate(uptake ~ Treatment + conc, data = CO2, FUN = summary)
```
