##### `ggplot` Plotting  

Now that we've covered `Base R` plotting, we'll look at plotting through `ggplot2` (sort of part of the tidyverse - [Cheat Sheet](https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=2&cad=rja&uact=8&ved=2ahUKEwjFu_ep-rroAhXnknIEHf13Bq4QFjABegQICBAB&url=https%3A%2F%2Frstudio.com%2Fwp-content%2Fuploads%2F2016%2F11%2Fggplot2-cheatsheet-2.1.pdf&usg=AOvVaw1-6jSR5VfWs7V62OYxtDtG)).  The main idea is outlined below:

- `ggplot(data = data_frame)` creates a plot instance  
- Add "layers" to the system (geoms or stats)  
    + Creates a visualization of the data  
- Modify layer "mapping" args (aes)
    + Ex: size, color, and x, y location(s)  
- Coordinate system (mostly use Cartesian plane)  
- Optional: Titles, etc.

We'll go through creation of the same plots in the `Base R` section.  

To create a barplot we can use `ggplot + geom_bar`.  To start, we create a base plotting object with `ggplot`.  Here we can specify attributes of the plot in the `aes` function.  For instance, to create a vertical bar chart we can specify the cateogrical variable of interest as `x`.  Let's do this for the `survived` variable from the titanic data set. `survived` is actually stored as a `dbl` or double in the dataset since it is coded as a 0/1 variable.  The plots will look better if we explicity change this as a `factor` first.  

```{r, eval=TRUE, out.width = "75%", fig.align='center'}
titanicData$survived <- as.factor(titanicData$survived)
ggplot(data = titanicData, aes(x = survived))
```

Notice that no plot is made!  We haven't added a `geom` or `stat` layer yet.  These are generally what you add in order to actually see something created with the plot.  The base plotting object just let's you specify global attributes like what goes on the x-axis.  Having the cheat sheet nearby is highly recommend when learning to use `ggplot`!  The `geom` layer we want to add is `geom_bar`.  

```{r,eval=TRUE, out.width = "75%", fig.align='center'}
ggplot(data = titanicData, aes(x = survived)) + geom_bar()
```

The usual practice is to save the base object and then add to it.  For instance:

```{r,eval = FALSE, out.width = "75%", fig.align='center'}
g <- ggplot(data = titanicData, aes(x = survived))
g + geom_bar()
```

We won't do this here just to help make the plots as clear as possible.  

You may wonder how to modify the plot.  `aes()` defines visual properties of objects in the plot.  This is where a lot of the modification occurs.  Changing these is similar to changing values of `par` in `Base R` plotting.  The cheat sheet gives the most commonly used aesthetics arguments for each `geom`.  Under `geom_bar` we see

> `d + geom_bar()`    
> `x, alpha, color, fill, linetype, size, weight`
    
The `data` and `aes` values can be set in two ways:

- 'globally' (for all layers) via the `ggplot` statement  

- 'locally' (for just that layer) via the `geom`, `stat`, etc. layer

```{r,eval = FALSE, out.width = "75%", fig.align='center'}
#global 
ggplot(data = titanicData, aes(x = survived)) + geom_bar()
#local 
ggplot() + geom_bar(data = titanicData, aes(x = survived))
```

To set an attribute that doesn't depend on the data (i.e. `color = 'blue'`), you generally place these outside of the `aes`.  Usually only aesthetics that need to depend on the data should go in the `aes` function.  

Back to our barplot.  We can improve the plot in a number of ways.  First, we can remove the `NA` (missing value) category by using the `drop_na` function.  Here we'll remove all rows where any of the three variables we are investigating have missing values.    

```{r,eval = TRUE, out.width = "75%", fig.align='center'}
titanicData <- titanicData %>% drop_na(survived, sex, embarked)
g <- ggplot(data = titanicData, aes(x = survived))
g + geom_bar()
```

Next, we can add better labels and a title by adding the appropriate layers (new layers --> see cheat sheet!).  The layers we want here are `labs` to set the title and `x` (or `y`) labels.  To change the names on the legend we need to look at the `scale_x_discrete` layer.  This is because the `aes` value's scale we are trying to modify is 'x' and we are doing so for a layer with a 'discrete' variable (their word for categorical).  

```{r,eval = TRUE, out.width = "75%", fig.align='center'}
ggplot(data = titanicData, aes(x = as.character(survived))) + 
  geom_bar() + 
  labs(x = "Survival Status", title = "Bar Plot of Survival for Titanic Passengers") + 
  scale_x_discrete(labels = c("Died", "Survived"))
```

A **stacked barplot** can be created by using the `fill` aesthetic and the same process.  
    
+ Create base object  
+ Add geoms  
+ Use aes to specify aspects of the plot  

The truly wonderful things about `ggplot` are the automatic assignment of colors, the creation of legends, and things of that nature for `aes` elements.  This takes a lot of work from the user.  Note: the `group` aestethic does not automatically generate a legend.  

```{r,eval = TRUE, out.width = "75%", fig.align='center'}
ggplot(data = titanicData, aes(x = survived, fill = sex)) + geom_bar()
```

Again, we can add custom labels by adding more layers.  

```{r,eval = TRUE, out.width = "75%", fig.align='center'}
ggplot(data = titanicData, aes(x = survived, fill = sex)) + 
  geom_bar() +
  labs(x = "Survival Status", 
       title = "Bar Plot of Survival Status for Titanic Passengers") + 
  scale_x_discrete(labels = c("Died", "Survived")) + 
  scale_fill_discrete(name = "Sex", labels = c("Female","Male"))
```

Notice the similar adjustment for the scale using `scale_*_discrete`.  For our aestethics, we set `aes(x = survived, fill = sex)`.  This means if we want to change one of those auto-generated scales we should use `scale_aes-value_discrete`.  

> `scale_x_discrete(labels = c("Died", "Survived"))`
> `scale_fill_discrete(name = "Sex", labels = c("Female","Male"))`

To turn the plot we can add the `coord_flip` layer.  

```{r,eval = TRUE, out.width = "75%", fig.align='center'}
ggplot(data = titanicData, aes(x = survived, fill = sex)) + geom_bar() +
  labs(x = "Survival Status", 
       title = "Bar Plot of Survival Status for Titanic Passengers") + 
  scale_x_discrete(labels = c("Died", "Survived")) + 
  scale_fill_discrete(name = "Sex", labels = c("Female","Male")) + 
  coord_flip()
```

One thing that can be quite confusing when you are first using the `ggplot` function is that most geoms have a corresponding stat that can be used instead.  For instance, look at the help for `geom_bar`:

`geom_bar(mapping = NULL, data = NULL, stat = "count",
  position = "stack", ..., width = NULL, binwidth = NULL, na.rm = FALSE,
  show.legend = NA, inherit.aes = TRUE)`

There is a `stat = "count"` default argument there.  An equivalent plot could be create using a `stat` layer instead.  

```{r,eval=FALSE,echo=TRUE}
ggplot(data = titanicData, aes(x = survived, fill = sex)) + geom_bar()
ggplot(data = titanicData, aes(x = survived, fill = sex)) + stat_count() 
```

Sometimes you'll need to modify the `stat`.  For example, if you have summary data to begin with rather than the raw data, change the stat to `identity`.  

```{r,eval = TRUE, out.width = "75%", fig.align='center'}
sumData <- titanicData %>% group_by(survived, sex) %>% summarize(count = n())
ggplot(sumData, aes(x = survived, y = count, fill = sex)) + 
  geom_bar(stat = "identity")
```

The **side-by-side barplot** can be created using the `position` aesthetic.  `position` has the following values:

+ `dodge` for side-by-side bar plot  
+ `jitter` for continuous data with many points at same values  
+ `fill` stacks bars and standardises each stack to have constant height  
+ `stack` stacks bars on top of each other

As the value of this aestethic doesn't depend on the data, we do not place this argument in the `aes` function.  

```{r,eval = TRUE, out.width = "75%", fig.align='center'}
ggplot(data = titanicData, aes(x = survived, fill = sex)) + 
  geom_bar(position = "dodge")
```

If we use `position = fill` the bars will be stacked and the bars will be standardized to have constant height.  This is really useful when there are an equal number of observations (or close to it) in each category on the x-axis (for instance with some Likert scale data surveys). 

```{r,eval = TRUE, out.width = "75%", fig.align='center'}
ggplot(data = titanicData, aes(x = survived, fill = sex)) + 
  geom_bar(position = "fill")
```

Where else does `ggplot` shine?  What if we want to make the same type of barplot for each `embarked` value?  Here we can use **faceting**!  This is just another layer on the plot call.  

`facet_wrap(~ var)` creates a plot for each setting of the variable `var`.  You can specify `nrow` and `ncol` or let R figure it out.  

```{r,eval = TRUE, out.width = "75%", fig.align='center'}
ggplot(data = titanicData, aes(x = survived)) + 
  geom_bar(aes(fill =sex), position = "dodge") +
  facet_wrap(~ embarked)
```

`facet_grid(var1 ~ var2)` creates a grid of plots for each combination of the variables `var1` and `var2`.  `var1` values go across rows and `var2` values go across columns.  To have side-by-side or a vertical column of plots use `. ~ var2` or `var1 ~ .`, respectively.  

```{r,eval = TRUE, out.width = "75%", fig.align='center'}
ggplot(data = titanicData, aes(x = survived)) + 
  geom_bar(aes(fill =sex), position = "dodge") +
  facet_wrap(~ embarked)
```

Here we'll create a grid of plots for all `pclass` and `embarked` combinations (so easy!).  
```{r,eval = TRUE, out.width = "75%", fig.align='center'}
ggplot(data = titanicData, aes(x = survived)) + 
  geom_bar(aes(fill =sex), position = "dodge") +
  facet_grid(pclass ~ embarked)
```

Below is a recap of creating a barplot with `ggplot`:

- Use `ggplot(data = ...) + geom_bar(aes(x = ...))`  
    
- With summary data, use `geom_bar(stat = identity)`  

- Stacked barplot via `aes(fill = )`

- Side-by-side barplot via `position = dodge`  

- Filled barplot via `position = fill`  

Recapping plotting with `ggplot` so far: 

- Can set local or global `aes`

- Modify titles/labels by adding more layers

- Use either stat or geom layer

- Faceting (multiple plots) via `facet_grid` or `facet_wrap`

- Only need `aes` if setting a mapping value that is dependent on the data (or you want to create a custom legend!)  


Now we'll create similar plots to the `Base R` section for our numeric variables from the `CO2` uptake dataset.  The process is the same:

- Create base object  
- Add geoms  
- Use aes to specify aspects of the plot  

Let's start by creating a histogram.  This can be done with `geom_bar`.  The easiest way to get started with a new `geom` is to look at the cheat sheet and the common `aes` values: 

> `c + geom_histogram(binwidth = 5)`
> `x, y, alpha, color, fill, linetype, size, weight`

For `geom_hist`, only `x` is really needed.  Now we'll create the histogram for the `uptake` variable.  

```{r,eval = TRUE, out.width = "75%", fig.align='center', message = FALSE, warning = FALSE}
g <- ggplot(CO2, aes(x = uptake))
g + geom_histogram()
```

We might modify a few attributes of the plot that are not dependent on the data.  Remember, these generally are set outside of the `aes` function.

```{r,eval = TRUE, out.width = "75%", fig.align='center'}
g + geom_histogram(color = "blue", fill = "red", size = 2, binwidth = 3)
```

To create the kernel smoother plot (smoothed histogram), we use the `geom_density` layer.  The common `aes` values from the cheat sheet are:

> `c + geom_density(kernel = "gaussian")`  

> `x, y, alpha, color, fill, group, linetype, size, weight`

As with `geom_hist`, only `x =` is really needed.  The kernel smoother for the `uptake` variable is created below.  

```{r, echo = TRUE, out.width = "75%", fig.align='center'}
ggplot(CO2, aes(x = uptake)) + geom_density()
```

This is not a pretty plot.  The `fill` aesthetic is usually used with `geom_density`.  Here, we'll fill by the `Treatment` variable.  We'll also change the `alpha` (transparency value, between 0 and 1) and the `adjust` (modify the kernel density calculations - 0.5 implies half the default bandwidth) aestethics.  

```{r, echo = TRUE, out.width = "75%", fig.align='center'}
ggplot(CO2, aes(x = uptake)) +
  geom_density(adjust = 0.5, alpha = 0.5, aes(fill = Treatment))
```

We may wish to stack the two density estimates on top of each other.  Recall the `position` aestethic and the choices of `dodge`, `jitter`, `fill`, and `stack`.

```{r, echo = TRUE, out.width = "75%", fig.align='center'}
ggplot(CO2, aes(x = uptake)) + 
  geom_density(adjust = 0.5, alpha = 0.5, position = "stack", aes(fill = Treatment))
```

Plots can easily be overlayed, as long as they are on the same scale, by simply adding the appropriate layers.  To put the histogram on the same scale as a density, we can set `y  = ..density..` on the `geom_hist` layer.  

```{r, echo = TRUE, out.width = "75%", fig.align='center', message = FALSE, warning = FALSE}
ggplot(CO2) + 
  geom_histogram(aes(y = ..density.., x = uptake, fill = Treatment)) +
  geom_density(adjust = 0.25, alpha = 0.5, position = "stack", 
               aes(x = uptake, fill = Treatment)) 
```

For these two plots, you can see it is a little beneficial to set global `aes()` options here.  This would allow for an easy change of the `fill` variable for all layers!

```{r,eval=FALSE,echo=TRUE,message=FALSE}
ggplot(CO2, aes(x = uptake, fill = Treatment)) + 
  geom_histogram(aes(y = ..density..)) +
  geom_density(adjust = 0.25, alpha = 0.5, position = "stack")
```

To create a boxplot we can use `geom_boxplot`.  The common `aes` values from the cheat sheet are:

> `f + geom_boxplot()`

> `x, y, lower, middle, upper, ymax, ymin, alpha, color, fill, group, linetype, shape, size, weight`

For `geom_boxplot`, only `x =, y =` are really needed.  To create a single vertical boxplot, use `x = ""`.  We'll create side-by-side boxplots of `uptake` for each `Treatment` level. 

```{r, echo = TRUE, out.width = "75%", fig.align='center'}
g <- ggplot(CO2, aes(x = Treatment, y = uptake))
g + geom_boxplot(fill = "grey")
```

Sometimes it is useful to overlay the data values themselves on a boxplot in order to see the shape of the distribution more clearly.  This can be done using the `geom_jitter` layer.  The `width` argument sets how widely the points are scattered (1 = width of box).

```{r, echo = TRUE, out.width = "75%", fig.align='center'}
ggplot(CO2, aes(x = Treatment, y = uptake)) + 
  geom_boxplot(fill = "grey") +
  geom_jitter(width = 0.1)
```

Note that the order of the layers is important!

```{r, echo = TRUE, out.width = "75%", fig.align='center'}
ggplot(CO2, aes(x = Treatment, y = uptake)) + 
  geom_jitter(width = 0.1) + 
  geom_boxplot(fill = "grey")
```

If we wanted to create these same side-by-side boxplots for each value of another variable, we can again use faceting.  For instance, we can use `facet_wrap` with the `Type` variable.  

```{r, echo = TRUE, out.width = "75%", fig.align='center'}
ggplot(CO2, aes(x = Treatment, y = uptake)) + geom_boxplot(fill = "grey") +
  geom_jitter(width = 0.1) + facet_wrap(~ Type)
```

Scatterplots can be created using `geom_point`.  The common `aes` values from the cheat sheet are:

> `e + geom_point()`
> `x, y, alpha, color, fill, shape, size, stroke`

For `geom_point`, only `x =, y =` are really needed.  

Below is a scatterplot of the `conc` and `uptake` variables. 

```{r, echo = TRUE, out.width = "75%", fig.align='center'}
g <- ggplot(CO2, aes(x = conc, y = uptake)) 
g + geom_point()
```

A trend lines can be easily added with `geom_smooth`.  The default is a smoothed line.  To request the simple linear regression fit, we can use `method = lm`.  Both of these lines are added to the plot below.  

```{r, echo = TRUE, out.width = "75%", fig.align='center', message = FALSE, warning = FALSE}
ggplot(CO2, aes(x = conc, y = uptake)) + geom_point() +  
  geom_smooth(aes(col = "loess")) +
  geom_smooth(method = lm, aes(col = "linear")) + 
  scale_colour_manual(name = 'Smoother', values = c('loess' = 'purple', 'linear' = 'red'), 
                      labels = c('GAM', 'Linear'), guide = 'legend')
```

You may notice something a bit odd in the call above.  The value of `col` (color) was set inside the `aes` function even though it was not set to a value corresponding to the data set.  This is an exception to that rule.  If you want to utilize `ggplot`'s automatic creation of legends, you can use the method above where labels are given to the color for each line ('loess' and 'linear', respectively).  `scale_colour_manual` is then used to create the legend.  Notice the `values` being set assign colors to the lines here.   

Text can also be added to the plot with `geom_text`.  The correlation is added to the plot in a similar way to how it was done in the `Base R` section.  The `size` argument being set to a value larger than 1 makes the text bigger.  

```{r, echo = TRUE, out.width = "75%", fig.align='center'}
correlation <- cor(CO2$conc, CO2$uptake)
ggplot(CO2, aes(x = conc, y = uptake)) + geom_point() +
  geom_smooth(method = lm, col = "Red") + 
  geom_text(x = 750, y = 10, size = 5, 
            label = paste0("Correlation = ", round(correlation, 2)))
```

Using text for the points is very easy with `geom_text`.  Just use this layer with a `label` variable.  

```{r, echo = TRUE, out.width = "75%", fig.align='center'}
ggplot(CO2, aes(x = conc, y = uptake)) + 
  geom_text(aes(label = Plant))
```

Again, if we want to create similar scatterplots for each value of another variable, we can do so with faceting.  (Note: the `cut` function is very useful for categorizing a numeric variable.)  Here we create similar scatterplots for each level of the `Treatment` variable.  

```{r, echo = TRUE, out.width = "75%", fig.align='center'}
ggplot(CO2, aes(x = conc, y = uptake)) + 
  geom_point(aes(color = Type), size = 2.5) +
  facet_wrap(~ Treatment)
```

Lastly, we saw the useful `pairs` function from `Base R`.  There are many extension packages for `ggplot` that give great functionality.  The `GGally` package has a function called `ggpairs` that is similar to pairs but retains the same syntax and functionality of `ggplot`.  Remember to install the package if you haven't done so and then read it in with the library command (or call the function directly using `::`).

```{r, echo = TRUE, out.width = "90%", fig.align='center', message = FALSE, warning = FALSE}
library(GGally)
ggpairs(iris, aes(colour = Species, alpha = 0.4))
```

To recap numerical variable plots using `ggplot`, the most common `geom`s are:

- Histogram (`geom_hist`), Density (`geom_density`)  

- Boxplot (`geom_boxplot`)

- Scatter plot (`geom_point`), Smoothers (`geom_smooth`)

- Jittered points (`geom_jitter`)  

- Text on plot (`geom_text`)  

The syntax and functionality of `ggplot` is recapped below again.

- Can set local or global `aes`

- Modify titles/labels by adding more layers

- Use either stat or geom layer

- Faceting (multiple plots) via `facet_grid` or `facet_wrap`

- Only need `aes` if setting a mapping value that is dependent on the data (or you want to create a custom legend!)  

- General syntax:  

`ggplot(data = <DATA>) +`  
&nbsp;&nbsp;`<GEOM_FUNCTION>(mapping = aes(<MAPPINGS>), stat = <STAT>, position = <POSITION>) +`  
&nbsp;&nbsp;`<COORDINATE_FUNCTION> +`  
&nbsp;&nbsp;`<FACET_FUNCTION> +`  
&nbsp;&nbsp;`<LABEL_FUNCTION>`   
`)`

There are many other useful extensions such as `gganimate`, `ggthemes`, `ggmap`, and `ggrepel`.
